Predicting prolonged apnea during nurse-administered procedural sedation using machine learning

### 2021-03-29 Extras/cover-letter.Rmd
submit jmir

---
date: "`r Sys.Date()`"
toaddress: Journal of Radiology Nursing
toname: Editorial Committee
output: letters::uhn_letter
knit: pagedown::chrome_print
---

@@

>>>>

---
date: "`r Sys.Date()`"
toaddress: JMIR Medical Informatics
toname: Editorial Committee
output: letters::uhn_letter
knit: pagedown::chrome_print
---

@@

I confirm that I agree to pay the article processing fee.

### 2021-03-29 manuscript.Rmd
submit jmir

---
title: "Predicting prolonged apnea during nurse-administered procedural sedation using machine learning"
bibliography: "Extras/references.bib"
csl: "Extras/anaesthesia.csl"
author:
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
  - Carla Jungquist PhD:
      institute: [UoB]
      correspondence: false
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Joanna Sutherland MBBS:
      institute: [NSW, Coffs]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - UoB: The University of Buffalo, Buffalo, New York, USA
  - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
  - Coffs: Department of Anaesthesia, Coffs Harbour Health Campus, Coffs Harbour, NSW, Australia
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesiology and Pain Medicine and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

Declarations of interest: none

Funding: A.C. received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S. Bloomberg Faculty of Nursing at the University of Toronto. This study received funding from the Wesley Medical Research (Project number: 2015-31).

Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.

@@

Capnography is commonly used for nurse-administered procedural sedation. This study aimed to determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will persist for 30 seconds or more. This information would help determine whether operationalizing these predictions in practice as alarm triggers would be beneficial.

@@

@@

Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementating this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates capnography's potential benefits on patient safety by improving ventilation. In practice, two alternative strategies for capnography alarm management are typically used. The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (e.g. 15 seconds). The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (e.g. 30 seconds).

It is possible that capnography monitor alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged. Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]

@@

Analyses were performed using R[@rcore2021]. Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.

#### Modelling
We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression, and the XGBoost model. Out of sample accuracy of the models was calculated using 10-fold cross-validation. Many participants in the study contributed multiple apneic events to the dataset used for modeling. To take this dependency into account, we ensured that apneic events from individual participants were not included in both the training and testing partitions of the 10-fold cross-validation process. Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and the total number of apneic events.

@@

A total of 384 apneic events of at least 15 seconds duration from 61 of the 102 patients who participated in this observational study were included in the present analysis. A summary of participant characteristics is presented in Table \@ref(tab:tab1). Nearly half of the (n=180) apneic events were prolonged, indicating a balanced dataset for the prediction problem at hand.

@@

The random forest had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03). A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value, and negative predictive value across all probability thresholds, is presented in Figure \@ref(fig:tpp).

@@

Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution for predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@krauss2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

@@

Conversely, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms. Using the random forest model as an additional input for an alarm trigger would reduce the total alarm burden and could be considered as an option for implementation into practice. To operationalize these predictions into capnography monitors, partnerships with industry would be required.[@glasgow2018nurse] Integrating predictive models into alarm management strategies for respiratory monitoring devices is also indicated in other contexts. For example, a recent study found that opioid-induced respiratory depression during recovery from anesthesia can be accurately predicted using a machine learning approach.[@jungquist2019identifying] In addition, user-centered design considerations, such as how the predictions should be communicated to nurses responsible for decision-making, would be important avenues for further research prior to implementation.[@risling2020advancing]

The finding that the model produced an overall net benefit against the aggressive capnography alarm management strategy but not the conservative approach is noteworthy. Further research with larger sample sizes is needed to increase the predictive power of models aimed at predicting the duration of apnea. Such research is warranted because triggering an alarm after 30 seconds of apnea that would have self-resolved without clinical intervention only 5 seconds later is just as clinically inconsequential as triggering an alert after 5 seconds of apnea that would similarly have resolved after a short period of time. In both of these circumstances, there would not have been enough time for the clinician's intervention to take effect. Yet, presumably in an attempt to reduce alarm burden, many capnography monitoring devices alarm settings default to triggering after apnea has been observed for 30 seconds. An ideal alternative to the conservative alarm strategy would be for clinicians to receive capnography monitor alarms as _early as possible_ in the course of an apneic event, _but only_ if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect. Results from the present study did not achieve this goal. Yet, previous research indicates that finding such a solution may be worthwhile. An analysis of over 500 thousand patients found that respiratory compromise in interventional radiology procedures performed with moderate sedation contributed to worse clinical outcomes and higher costs.[@urman2019impact]

@@

>>>>

---
title: "Predicting prolonged apnea during nurse-administered procedural sedation using machine learning"
bibliography: "Extras/references.bib"
csl: "Extras/jmir.csl"
author:
  - Aaron Conway PhD:
      email: aaron.conway@utoronto.ca
      phone: +1 416 946 7112
      address: 585 University Ave | Toronto, ON  M5G 2N2
      institute: [PMCC, LBFON, QUT]
      correspondence: true
  - Carla Jungquist PhD:
      institute: [UoB]
      correspondence: false
  - Kristina Chang MScN:
      institute: [PMCC]
      correspondence: false
  - Navpreet Kamboj BScN:
      institute: [LBFON]
      correspondence: false
  - Joanna Sutherland MBBS:
      institute: [NSW]
      correspondence: false
  - Sebastian Mafeld:
      institute: [IR]
      correspondence: false
  - Matteo Parotto MD, PhD:
      institute: [TGH, UoT]
      correspondence: false
institute:
  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
  - UoB: The University of Buffalo, Buffalo, New York, USA
  - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
  - IR: Interventional Radiology, JDMI, Toronto General Hospital
  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
  - UoT: Department of Anesthesiology and Pain Medicine and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
output:
  officedown::rdocx_document:
    reference_docx: Extras/manuscript_template.docx
    pandoc_args:
      - --lua-filter=Extras/scholarly-metadata.lua
      - --lua-filter=Extras/author-info-blocks.lua
---

@@

### Background

Capnography is commonly used for nurse-administered procedural sedation. Deciphering which capnography waveform abnormalities deserve intervention, and therefore alarms to signal the event to clinicians, from those that do not is an essential step towards successfully implementing this technology into practice. It is possible that capnography alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.

### Objective

To determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will be prolonged (defined as apnea that persisted for 30 seconds or more).

### Methods

@@

**Keywords**

Procedural sedation and analgesia; conscious sedation; nursing; informatics; patient safety; machine learning; capnography; anesthesia

@@

Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementing this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates capnography's potential benefits on patient safety by improving ventilation. In practice, two alternative strategies for capnography alarm management are typically used. The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (e.g. 15 seconds). The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (e.g. 30 seconds). Preferences for the aggressive or conservative alarm threshold will be influenced by many factors, including the rate of oxygen supplmentation. The duration of time between the onset of apnea to hypoxemia increases with higher flows of oxygen.[@wong2019high]

It is possible that capnography alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged. Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]

@@

Analyses were performed using R version 4.0.3.[@rcore2021] Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available.[@conway2021code; @conway2021data]

#### Modelling

We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression, and the XGBoost model. Out of sample accuracy of the models was calculated using 10-fold cross-validation. Many participants in the study contributed multiple apneic events to the dataset used for modeling. To take this dependency into account, we ensured that apneic events from individual participants were not included in both the training and testing partitions of the 10-fold cross-validation process. Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and the total number of apneic events. Discriminative ability of the models was compared using the area under receiver operating characteristics curve (AUROC) as well as by plotting sensitivity, specificity, positive predictive values and negative predictive values (termed a threshold performance plot). A calibration plot with a loess smoother was used to assess calibration.[@van2016calibration] The `runway` package was used to create the plots.[@singh2021]

@@

A total of 384 apneic events of at least 15 seconds duration from 61 of the 102 patients who participated in this observational study were included in the present analysis. A summary of participant characteristics is presented in Table \@ref(tab:tab1). Nearly half of the (n=180) apneic events were prolonged (i.e. >30 seconds), indicating a balanced dataset for the prediction problem at hand.

@@

The random forest had the best discriminatory power of the models, with a mean AUROC score of 0.66 (standard error 0.03). A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value, and negative predictive value across all probability thresholds, is presented in Figure \@ref(fig:tpp).

@@

Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution for predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (hazard ratio 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (hazard ratio 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@krauss2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

@@

Conversely, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms. Using the random forest model as an additional input for an alarm trigger would reduce the total alarm burden and could be considered as an option for implementation into practice. To operationalize these predictions into capnography monitors, partnerships with industry would be required because monitor functionality would need to be adapted to facilitate input of the data required to calculate the predictions.[@glasgow2018nurse] These would include patient characteristics and sedation dosing. Integrating predictive models into alarm management strategies for respiratory monitoring devices is also indicated in other contexts. For example, a recent study found that opioid-induced respiratory depression during recovery from anesthesia can be accurately predicted using a machine learning approach.[@jungquist2019identifying] In addition, user-centered design considerations, such as how the predictions should be communicated to nurses responsible for decision-making, would be important avenues for further research prior to implementation.[@risling2020advancing]

The finding that the model produced an overall net benefit against the aggressive capnography alarm management strategy but not the conservative approach is noteworthy. Further research with larger sample sizes is needed to increase the predictive power of models aimed at predicting the duration of apnea. Such research is warranted because triggering an alarm after 30 seconds of apnea that would have self-resolved without clinical intervention only 5 seconds later is just as clinically inconsequential as triggering an alert after 5 seconds of apnea that would similarly have resolved after a short period of time. In both of these circumstances, there would not have been enough time for the clinician's intervention to take effect. Yet, presumably in an attempt to reduce alarm burden, many capnography monitoring devices alarm settings default to triggering after apnea has been observed for 30 seconds. An ideal alternative to the conservative alarm strategy would be for clinicians to receive capnography monitor alarms as _early as possible_ in the course of an apneic event, _but only_ if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect. Results from the present study did not achieve this goal. Yet, previous research indicates that finding such a solution may be worthwhile. An analysis of half a million patients found that respiratory compromise in interventional radiology procedures performed with moderate sedation contributed to worse clinical outcomes and higher costs.[@urman2019impact]

@@

## Ackowledgements

A.C. received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S. Bloomberg Faculty of Nursing at the University of Toronto. This study received funding from the Wesley Medical Research (Project number: 2015-31).
Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.

## Conflicts of interest

None

## Abbreviations

AUROC: Area under the receiver operating characteristics curve
ETCO~2~: End-tidal carbon dioxide
CO~2~: Carbon dioxide
